State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2021-05-06 analysis/paper/paper.Rmd
changed some latex formatting so equations are numbered. responded to a few very minor requested revisions from associate editor

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon, *Oncorhynchus tshawytscha*, and steelhead, *O. mykiss*, from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

$$
\begin{aligned}
  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
  e_t &\sim \mathcal{N}(0, \sigma^2_X)
\end{aligned}
$$

@@

$$
Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right)
$$

@@

$$
m_t \sim \text{Bin} \left( \nu_t, M_t \right)
$$

@@

$$
\begin{aligned}
  X_t^{day} &= X_t * \theta_t \\
  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
\end{aligned}
$$

@@

$$
\begin{aligned}
  y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
  \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
  g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
\end{aligned}
$$

@@

$$
\begin{aligned}
  y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
  \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
  f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
\end{aligned}
$$

@@

After estimating the total number of fish to have crossed LGR each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$. This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$. The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$. The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time, while smoothing over large weekly fluctuations in the proportions due to small sample size. This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap. The random walk deviations, $d_{w,t}$ and $d_{hnc,t}$, are drawn from the same normal distribution with variance $\sigma^2_\omega$.

$$
\begin{aligned}
  \left( Y^T_{w,t}, Y^T_{hc,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
  \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
  \phi_{hc,t} &= 0 \\
  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{hc,t}} \right) \\
  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{hc,t}} \right) \\
  \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
  \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
  d_{w,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
  d_{hnc,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
\end{aligned}
$$

@@

$$
\begin{aligned}
  \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
\end{aligned}
$$
In practice, the total number of fish crossing LGR each week, $X_t$, could be split into other discrete groups if desired; for example, individual wild stocks using genetic stock identification or hatchery stocks with parentage based tagging [@Steele2019].

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon, *Oncorhynchus tshawytscha*, and steelhead, *O. mykiss*, from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

\begin{equation}
  \begin{aligned}
    \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
    e_t &\sim \mathcal{N}(0, \sigma^2_X)
  \end{aligned}
\end{equation}

@@

\begin{equation}
  Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right)
\end{equation}

@@

\begin{equation}
  m_t \sim \text{Bin} \left( \nu_t, M_t \right)
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    X_t^{day} &= X_t * \theta_t \\
    p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\  
    Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
  \end{aligned}
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
    \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
    g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
  \end{aligned}
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
    \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
    f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
  \end{aligned}
\end{equation}

@@

After estimating the total number of fish to have crossed LGR each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$. This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$. The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$. The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish $\omega_{hc,t}$ ($\phi_{w,t}$ and $\phi_{hnc,t}$), is modeled as a random walk, so it can change through time, while smoothing over large weekly fluctuations in the proportions due to small sample size. This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap. The random walk deviations, $d_{w,t}$ and $d_{hnc,t}$, are drawn from the same normal distribution with variance $\sigma^2_\omega$.

\begin{equation}
  \begin{aligned}
    \left( Y^T_{w,t}, Y^T_{hc,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
    \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
    \phi_{hc,t} &= 0 \\
    \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{hc,t}} \right) \\
    \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{hc,t}} \right) \\
    \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
    \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
    d_{w,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
    d_{hnc,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
  \end{aligned}
\end{equation}

@@

\begin{equation}
  \begin{aligned}
    \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
    &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
  \end{aligned}
\end{equation}

In practice, the total number of fish crossing LGR each week, $X_t$, could be split into other discrete groups if desired; for example, individual wild stocks using genetic stock identification or hatchery stocks with parentage based tagging [@Steele2019]. A graphical depiction (directed acyclic graph) of the model framework described above is found in Figure \@ref(fig:dag-fig).
